Cluster-efficient pangenome graph construction with nf-core/pangenome

### 2024-10-09 manuscript/main.tex
references: remove URLs, added to main text

	\subsection{Pipeline overview}
	The pipeline’s (Fig. \ref{fig:workflow}) input is a FASTA file compressed with \textit{bgzip} \citep{Li2009} containing the sequences to create the graph. 
	Sequence names should follow the Pangenome Sequence Naming specification (PanSN-spec) \citep{pansn-spec}. 
	The primary output is a pangenome variation graph \citep{Garrison:2018} in the Graphical Fragment Assembly (GFA) format version 1 \citep{GFA}.

@@

	The pipeline begins with an all-to-all alignment of the input sequences using the whole-chromosome pairwise sequence aligner WFMASH \citep{wfmash}, avoiding reference, order, or orientation bias, allowing every sequence to serve as a reference.
	%This avoids reference, order, or orientation bias, and allows each sequence in the pangenome to serve as a reference when exploring related variation. 
	In the pangenome graph induction step SEQWISH \citep{Garrison2022}, an alignment to variation graph inducer, converts the sequence alignments into a variation graph. 
	%SEQWISH can recapture homology relationships that may not be encoded in the initial alignments. 
	%This allows us to randomly sparsify the mappings computed by WFMASH. 
	%The applied heuristic is based on the Erdös-Rényi random graph model \citep{Bollobs2001}. 
	The graph is then simplified using SMOOTHXG \citep{Garrison2023}:
	A 1-dimensional (1D) graph embedding \citep{Heumos2024} orders the graph’s nodes to best match the nucleotide distances of the genomic paths of the graph.
	%This considers the local spurious complexity inherent in the raw pairwise alignments. 
	%If ignored, this could lead to difficulties in downstream processing of the graph.
	%Therefore, SMOOTHXG iteratively applies a local Multiple Sequence Alignment (MSA), partial order alignment (POA) \citep{Lee2002}, to refine and compress the pangenome graph. A 1-dimensional (1D) graph embedding \citep{Heumos2023} gives rise to the regions the kernel is applied to. 
	%After applying the graph embedding algorithm, the graphs’ node order best-matches the nucleotide distances of the genomic paths of the graph. 
	Next, the graph is split into partially overlapping segments. 
	%The sequences of each segment are realigned with POA. 
	The sequences of each segment are realigned with a local Multiple Sequence Alignment (MSA) kernel, partial order alignment (POA) \citep{Lee2002}.
	Afterwards, the segments are laced back together into a variation graph. 
	By default, the SMOOTHXG process is applied 3 times in order to smooth the edge effects at the boundaries of the segments.
	Finally, we employ GFAFFIX \citep{Liao2023} to systematically condense redundant nodes within the graph.
	%In the final normalization phase, we employ GFAFFIX \citep{Liao2023} to systematically condense redundant nodes within the graph. 
	%Subsequently, the 1D graph embedding algorithm is reapplied to ensure a human-readable representation of the processed data. 
	%For more details about each step, we refer to the original PGGB paper (Garrison et al. 2023).

@@

	Although we expect our pipeline to scale for future challenges, such as for the next HPRC phase which targets 350 individuals, further optimizations are possible: 
	The IMplicit Pangenome Graph (IMPG) (\href{https://github.com/ekg/impg}{https://github.com/ekg/impg}) tool extracts homologous loci from genomes mapped to a specific target region.
	This would allow us to break the whole genome multiple alignments into smaller pieces, construct a pangenome graph for each piece, and lace these together into a full graph with \href{https://github.com/pangenome/gfalace}{https://github.com/pangenome/gfalace}. 
	%Our first trials 
	%(\href{https://hackmd.io/vh0Ap65uQTm40DdleA-Oow}{https://hackmd.io/vh0Ap65uQTm40DdleA-Oow}) 
	%suggest following this direction.

>>>>

	\subsection{Pipeline overview}
	The pipeline’s (Fig. \ref{fig:workflow}) input is a FASTA file compressed with \textit{bgzip} \citep{Li2009} containing the sequences to create the graph. 
	Sequence names should follow the Pangenome Sequence Naming specification (PanSN-spec) (\href{https://github.com/pangenome/PanSN-spec}{https://github.com/pangenome/PanSN-spec}, last accessed October 2024). 
	The primary output is a pangenome variation graph \citep{Garrison:2018} in the Graphical Fragment Assembly (GFA) format version 1 (\href{http://gfa-spec.github.io/GFA-spec/GFA1.html}{http://gfa-spec.github.io/GFA-spec/GFA1.html}, last accessed October 2024).

@@

	The pipeline begins with an all-to-all alignment of the input sequences using the whole-chromosome pairwise sequence aligner WFMASH (\href{https://github.com/waveygang/wfmash}{https://github.com/waveygang/wfmash}, last accessed October 2024), avoiding reference, order, or orientation bias, allowing every sequence to serve as a reference.
	%This avoids reference, order, or orientation bias, and allows each sequence in the pangenome to serve as a reference when exploring related variation. 
	In the pangenome graph induction step SEQWISH \citep{Garrison2022}, an alignment to variation graph inducer, converts the sequence alignments into a variation graph. 
	%SEQWISH can recapture homology relationships that may not be encoded in the initial alignments. 
	%This allows us to randomly sparsify the mappings computed by WFMASH. 
	%The applied heuristic is based on the Erdös-Rényi random graph model \citep{Bollobs2001}. 
	The graph is then simplified using SMOOTHXG \citep{Garrison2023}:
	A 1-dimensional (1D) graph embedding \citep{Heumos2024} orders the graph’s nodes to best match the nucleotide distances of the genomic paths of the graph.
	%This considers the local spurious complexity inherent in the raw pairwise alignments. 
	%If ignored, this could lead to difficulties in downstream processing of the graph.
	%Therefore, SMOOTHXG iteratively applies a local Multiple Sequence Alignment (MSA), partial order alignment (POA) \citep{Lee2002}, to refine and compress the pangenome graph. A 1-dimensional (1D) graph embedding \citep{Heumos2023} gives rise to the regions the kernel is applied to. 
	%After applying the graph embedding algorithm, the graphs’ node order best-matches the nucleotide distances of the genomic paths of the graph. 
	Next, the graph is split into partially overlapping segments. 
	%The sequences of each segment are realigned with POA. 
	The sequences of each segment are realigned with a local Multiple Sequence Alignment (MSA) kernel, partial order alignment (POA) \citep{Lee2002}.
	Afterwards, the segments are laced back together into a variation graph. 
	By default, the SMOOTHXG process is applied 3 times in order to smooth the edge effects at the boundaries of the segments.
	Finally, we employ GFAFFIX \citep{Liao2023} to systematically condense redundant nodes within the graph.
	%In the final normalization phase, we employ GFAFFIX \citep{Liao2023} to systematically condense redundant nodes within the graph. 
	%Subsequently, the 1D graph embedding algorithm is reapplied to ensure a human-readable representation of the processed data. 
	%For more details about each step, we refer to the original PGGB paper (Garrison et al. 2023).

@@

	Although we expect our pipeline to scale for future challenges, such as for the next HPRC phase which targets 350 individuals, further optimizations are possible: 
	The IMplicit Pangenome Graph (IMPG) (\href{https://github.com/ekg/impg}{https://github.com/ekg/impg}, last accessed October 2024) tool extracts homologous loci from genomes mapped to a specific target region.
	This would allow us to break the whole genome multiple alignments into smaller pieces, construct a pangenome graph for each piece, and lace these together into a full graph with \href{https://github.com/pangenome/gfalace}{https://github.com/pangenome/gfalace}. 
	%Our first trials 
	%(\href{https://hackmd.io/vh0Ap65uQTm40DdleA-Oow}{https://hackmd.io/vh0Ap65uQTm40DdleA-Oow}) 
	%suggest following this direction.
